---
title: "Untitled"
output: html_document
date: "2025-09-05"
---


  * Maisie shared an updated version of the GPS coordinates CSV, including the date and time of each assessment. The tasks are to 
  
    - generate NDVI for each coordinate within the month of the assessment date, using a 100m buffer
    - generate NDVI for each coordinate within the month of the assessment date, using a 200m buffer 
    - generate NDVI for each coordinate within the month of the assessment date, using a 300m buffer 
    - generate NDVI for each coordinate matched to the season and year, using a 300m buffer 
    
    
  * Steps: 

    - create buffer zones and save as polygons
    - upload polygons to GEE to extract NDVI values
  

```{r include=FALSE}

library(readr)
library(dplyr)
library(sf)
library(lubridate)
library(writexl)
library(ggplot2)

library(purrr)
library(tidyr)
library(stringr)

ymd <- format(Sys.time(), "%Y%m%d"); ymd
```



## Load data and pre-cleaning

```{r social mind data}

dir <- "G:/Shared drives/Wellcome Trust Project Data/0_source_data/"


```


```{r}

## 1. load NDVI data from GEE
dir.ndvi <- 'D:/SocialMind-NDVI-coords'

# ## test
# ndvi <- read_csv(file = file.path(dir.ndvi, 'sm_coords_ndvi_2020_Fall_200m.csv'), show_col_types = F) 
# names(ndvi)



## 1.2 load all


all_files <- list.files(dir.ndvi, pattern = "^sm_coords_ndvi.*\\.csv$", full.names = TRUE)

# ---------------------------
# Seasonal: Spring, Summer, Fall, Winter
# ---------------------------
seasonal_files <- all_files[grepl("Spring|Summer|Fall|Winter",
                                  basename(all_files))]

# ---------------------------
# Monthly: Jan, Feb, Mar ... Dec
# ---------------------------
months <- paste(month.abb, collapse = "|")   # "Jan|Feb|Mar|Apr|May|Jun|Jul|Aug|Sep|Oct|Nov|Dec"

monthly_files <- all_files[grepl(months, basename(all_files))]

# ---------------------------
# Check
# ---------------------------
message("Seasonal files: ", length(seasonal_files)) # season * year * buffer =  4*5*3 = 60
message("Monthly files:  ", length(monthly_files))  # month  * year * buffer = 12*5*3 = 180

# basename(seasonal_files)
# basename(monthly_files)
```


```{r - season}

ndvi.s <- seasonal_files %>%
  map_dfr(~ read_csv(.x, show_col_types = FALSE) %>%
            select(uid, everything())
          ) %>%
  mutate(adate = as.Date(as.POSIXct(adate/1000, origin = "1970-01-01", tz = "UTC")) )


## test on one example ---
# ndvi.s.i <- ndvi.s %>%
#   filter(uid == 2) %>% 
#   # filter(buffer_m == 200) %>%
#   # match date and season 
#   filter(adate >= date_ini & adate <= date_end) %>%
#   # if duplicates exist, keep the season with the closest date_ini
#   group_by(uid, buffer_m) %>%
#   slice_min(abs(as.numeric(adate - date_ini))) %>%
#   ungroup() %>%
#   as.data.frame()




ndvi_season <- ndvi.s %>%
  # match date and season 
  filter(adate >= date_ini & adate <= date_end) %>%
  # if duplicates exist, keep the season with the closest date_ini
  group_by(uid, buffer_m) %>%
  slice_min(abs(as.numeric(adate - date_ini))) %>%
  ungroup() %>%
  as.data.frame()
  
length(unique(ndvi_season$uid))

4540*3
```


```{r - month}

ndvi.m <- monthly_files %>%
  map_dfr(~ read_csv(.x, show_col_types = FALSE) %>%
            select(uid, everything())
          ) %>%
  mutate(adate = as.Date(as.POSIXct(adate/1000, origin = "1970-01-01", tz = "UTC")) )

ndvi_month <- ndvi.m %>%
  # match date and season 
  filter(adate >= date_ini & adate <= date_end) %>%
  # if duplicates exist, keep the season with the closest date_ini
  group_by(uid, buffer_m) %>%
  slice_min(abs(as.numeric(adate - date_ini))) %>%
  ungroup() %>%
  select(-month_name) %>%
  as.data.frame()


length(unique(ndvi_month$uid))


nrow(sm_sf_m)*12*3

```


```{r - save}

## 3. save and share with collaborator 

## - season
f <- file.path(dir, 'SocialMind', 'output_coords', paste0('sm_coords_NDVI_season_', ymd, '.csv')); f
readr::write_csv(x = ndvi_season, file = f)

## - monthly
f <- file.path(dir, 'SocialMind', 'output_coords', paste0('sm_coords_NDVI_monthly_', ymd, '.csv')); f
readr::write_csv(x = ndvi_month, file = f)

union(names(ndvi_season), names(ndvi_month))

```



