---
title: "Untitled"
output: html_document
date: "2025-09-05"
---


  * Maisie shared an updated version of the GPS coordinates CSV, including the date and time of each assessment. The tasks are to 
  
    - generate NDVI for each coordinate within the month of the assessment date, using a 100m buffer
    - generate NDVI for each coordinate within the month of the assessment date, using a 200m buffer 
    - generate NDVI for each coordinate within the month of the assessment date, using a 300m buffer 
    - generate NDVI for each coordinate matched to the season and year, using a 300m buffer 
    
    
  * Steps: 

    - create buffer zones and save as polygons
    - upload polygons to GEE to extract NDVI values
  

```{r include=FALSE}

## To clear your environment
remove(list = ls())

library(readr)
library(dplyr)
library(sf)
library(lubridate)
library(writexl)
library(ggplot2)

library(purrr)
library(tidyr)
library(stringr)

ymd <- format(Sys.time(), "%Y%m%d"); ymd
```



## Load data and pre-cleaning

```{r social mind data}

dir <- "G:/Shared drives/Wellcome Trust Project Data/0_source_data/"

dir.sm <- file.path(dir, 'SocialMind', 'UPDATED GPS coordinates Social Mind.csv') # 3/20/2026
dir.postcode <- file.path(dir, 'Geolocation Data', 'london_postcode.csv')

sm_raw <- read_csv(dir.sm, show_col_types = F) %>%
  rename('lat' = 'startlatitude',
         'lon' = 'startlongitude') %>%
  mutate(adate = as.Date(actualstarttime)) %>%
  mutate(uid = row.names(.)) 

names(sm_raw)

sm_ <- sm_raw %>%
  filter(!is.na(lat)) %>%
  filter(!is.na(lon)) %>%
  select(-actualstarttime) %>%
  rename('pid' = 'participantid',
         'aid' = 'assessmentnumber') %>%
  as.data.frame()


sm <- sm_ %>%
  filter(lat > 50 & lat < 53) %>%
  filter(lon > -2 & lon < 2) %>%
  as.data.frame()

names(sm)
```


```{r summary}

# ---------------------------
# Summary: missing values + basic stats
# ---------------------------
sm_summary_missing <- sm_raw %>%
  summarise(across(everything(), list(
    n_total   = ~n(),
    n_missing = ~sum(is.na(.)),
    pct_missing = ~round(mean(is.na(.)) * 100, 1)
  ))) %>%
  pivot_longer(
    cols      = everything(),
    names_to  = c("variable", ".value"),
    names_pattern = "^(.+)_(n_total|n_missing|pct_missing)$"  # ← explicit suffixes
    # names_sep = "_(?=[^_]+$)"          # split on last underscore
  ) %>%
  mutate(n_complete = n_total - n_missing) %>%
  select(variable, n_total, n_complete, n_missing, pct_missing)


# ---------------------------
# Save
# ---------------------------
f_summary <- gsub("\\.csv$", "_data_summary.csv", dir.sm)   # saves alongside raw file
write_csv(sm_summary_missing, f_summary)
message("Saved: ", f_summary)
```




```{r df to sf}
# Convert to sf
sm_sf <- sm %>%
  st_as_sf(., coords = c("lon", "lat"), crs = 4326, remove = FALSE)  # WGS84

# plot(sm_sf[1])

## save data - only shp format can be uploaded to GEE
f <- file.path(dir, 'SocialMind', 'output_coords', paste0('sm_coords.shp')); f
st_write(sm_sf, f, delete_layer = TRUE)


# 1. Reproject to a CRS in meters (here: British National Grid, EPSG:27700)
sm_sf_m <- st_transform(sm_sf, 27700)

```


```{r sf with buffer, eval=FALSE, include=FALSE}

# 2. Create x m buffer
buffer_m <- 100
# buffer_m <- 200
# buffer_m <- 300

buffers <- st_buffer(sm_sf_m, dist = buffer_m) %>%
  mutate(buffer_m = buffer_m)

# 3. (Optional) Transform back to WGS84 for mapping
##'   Warning: All vector data in Earth Engine is stored in a spherical coordinate system (specifically, EPSG:4326).To minimize the possibility of unintended changes, first reproject vector data to EPSG:4326 before uploading.
buffers_wgs84 <- st_transform(buffers, 4326)


## save data
## --> only shp format can be uploaded to GEE
f <- file.path(dir, 'SocialMind', 'output_coords', paste0('sm_coords_', buffer_m, 'm.shp')); f
st_write(buffers_wgs84, f, delete_layer = TRUE)

```

```{r batch buffer}

library(purrr)

# 2. Create buffers for all sizes, transform, and save in one pass
buffer_sizes <- c(100, 200, 300)

walk(buffer_sizes, function(buffer_m) {
  
  f <- file.path(dir, 'SocialMind', 'output_coords',
                 paste0('sm_coords_', buffer_m, 'm.shp'))
  
  sm_sf_m %>%
    st_buffer(dist = buffer_m) %>%
    mutate(buffer_m = buffer_m) %>%
    st_transform(4326) %>%                        # reproject to WGS84 for GEE
    st_write(f, delete_layer = TRUE)
  
  message("Saved: ", basename(f))
})

```





The seasonal division is consistent with Yougeng’s earlier approach

  Spring: 3/1 - 6/1
  Summer: 6/1 - 9/1
  Fall:   9/1 - 12/1
  Winter: 12/1 - 3/1
  
  
## For GEE

Then upload the shapefile data to GEE, and use GEE to calculate NDVI by postcode region 

The GEE script name is `london_urban_cooling/SocialMind_NDVI_coords`

```{r}

```


