---
title: "Figure 4a — citywide avoided energy cost"
output: html_document
date: "2026-09-11"
---

## Production panel (nine manuscript scenarios)

This section is the current manuscript workflow. It uses the reviewed citywide
summary, in which every building is counted once. It retains AllBuilt, TreeRisk
and TreeOpp and displays Green and Target as matched 10%, 20% and 30% pairs.
Set `URBAN_COOLING_DATA_ROOT` to
the local **Wellcome Trust Project Data** folder before knitting.

```{r production-setup, message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
library(here)

source(here("code", "post_processing_layers", "figure4_panel_helpers.R"))

data_root <- figure4_data_root()
valuation_root <- file.path(
  data_root, "2_postprocess_intermediate", "UCM_official_runs",
  "revised_equal_area_invest3202_2026-09-11_valuations"
)
summary_path <- file.path(
  valuation_root, "summary", "citywide_energy_productivity_summary.csv"
)
output_dir <- figure4_output_dir(
  file.path(valuation_root, "summary", "figure4", "source_panels")
)
temperature_c <- 25
stopifnot(file.exists(summary_path))
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)
```

```{r production-energy-data}
# `energy_change_vs_baseline` is already scenario minus baseline under the
# matching 25°C climate condition; convert pounds to £ million per month.
energy_panel_data <- read.csv(summary_path, stringsAsFactors = FALSE) %>%
  filter(temperature_c == .env$temperature_c, scenario %in% figure4_scenarios) %>%
  figure4_prepare_scenarios() %>%
  transmute(
    scenario_id, scenario, strategy,
    estimate = energy_change_vs_baseline / 1e6
  )

write.csv(
  energy_panel_data,
  file.path(output_dir, "figure4a_energy_citywide_data.csv"),
  row.names = FALSE
)
```

```{r production-energy-plot, fig.width=5.2, fig.height=4.2}
figure4a <- ggplot(
  energy_panel_data, aes(x = scenario, y = estimate, fill = strategy)
) +
  geom_hline(yintercept = 0, colour = "#777777", linewidth = 0.3) +
  geom_vline(xintercept = 3.5, colour = "#B8B8B8", linewidth = 0.35) +
  geom_col(width = 0.72, colour = "#2B2B2B", linewidth = 0.25) +
  scale_fill_manual(values = figure4_colours) +
  scale_y_continuous(expand = expansion(mult = c(0.08, 0.10))) +
  labs(
    x = NULL, y = "Avoided energy cost (£ million/month)",
    fill = "Scenario strategy"
  ) +
  figure4_panel_theme()

figure4a
figure4_save_panel(figure4a, output_dir, "figure4a_energy_citywide")
```

The plotted quantity is a citywide total across unique buildings relative to
baseline. Energy prices use late-2025 input-price assumptions.

## Archived manuscript-era workflow

The code below is retained to document the former equal-weight borough
workflow and its historical inputs. It is not executed when this notebook is
knit; use the production section above for the revised manuscript panel.

```{r archive-options, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

-->
--> Note: run the `invest_result_zonal_viz_1_temp.Rmd` first, before run this script. 
--> 


  Re-use the paths and functions in the script

## Packages and functions 

```{r}
# load functions 
f <- here("code", "func_calculate_change_from_yri_lc0.rds");
func_calculate_change_from_yri_lc0 <- readRDS(f)

f <- here('code', 'func_calculate_change_from_yr0_lc0.rds');
func_calculate_change_from_yr0_lc0 <- readRDS(f)

f <- here('code', 'func_combine_two_change_dfs.rds')
func_combine_two_change_dfs <- readRDS(f) 

f <- here('code', 'func_plot_line.rds')
func_plot_line <- readRDS(f)

f <- here('code', 'func_plot_change_point.rds')
func_plot_change_point <- readRDS(f)

f <- here('code', 'func_plot_map.rds')
func_plot_map <- readRDS(f)
```


## AOI

```{r - borough}

#' borough level
f <- file.path(dir.aoi, 'London_Borough_aoi.shp')
aoi <- st_read(f, quiet = T) %>% select(GSS_CODE)

```


```{r - lsoa}

#' lsoa data with quintiles info, which was generated by `socio-economic-data-ES-equity.Rmd`
f <- './data/Social_Vulnerability_Index_london_q.gpkg'
aoi_lsoa <- st_read(f, quiet = T) %>%
  rename('GSS_CODE' = 'id') %>%
  mutate(GSS_CODE = as.character(GSS_CODE)) %>%
  select(GSS_CODE)

```


## Energy saving - select one input

  `avd_eng_cn` - (optional) Avoided energy consumption (kWh or $ if optional energy cost input column was provided in the Energy Consumption Table).

### ? Pick one scale ---

### Load data - borough level 

```{r}

sp_scale = "borough"
ind <- 'energy_sav_sum' # --> sum or average
ind_scale <- 1/10^6  # convert to million
ind_label <- "avoided energy cost (Million £)"


#' load data generated from `invest_result_zonal_viz_0_data_prep.Rmd`
# f <- paste0('./data/', 'ucm_output_shp_all_df.rds'); f # old name
f <- paste0('./data/', 'ucm_output_shp_stats_outcomes_by_borough.rds'); f # new name
shp_all_df <- readRDS(f) %>%
  rename(ind = 'avd_eng_cn')

# 1. calculate the change from year i
df_stats_i0 <- func_calculate_change_from_yri_lc0(data = shp_all_df, aoi = aoi)
df_chg_sf_i0<- func_calculate_change_from_yri_lc0(data = shp_all_df, aoi = aoi, return_sf = T)


# 2. also calculate the change from year 0
df_stats_00 <- func_calculate_change_from_yr0_lc0(data = shp_all_df, aoi = aoi)

# 3. and combine the two change calculations  
stat_both_l <- func_combine_two_change_dfs(df_stats_00, df_stats_i0)

```


### Load data - lsoa level 

```{r eval=FALSE, include=FALSE}
sp_scale = "lsoa"
ind <- "energy_sav_sum"; 
# ind <- "energy_sav_mean"
ind_scale <- 1/10^6  # convert to million
ind_label <- "avoided energy cost (Million £)"

f <- paste0('./data/', 'ucm_output_shp_stats_energy_by_lsoa.rds'); f
shp_all_df <- readRDS(f) %>%
  rename('GSS_CODE' = 'LSOA_CODE') %>% # rename to be able to use the function 
  mutate(GSS_CODE = as.character(GSS_CODE))

# 1. calculate the change from year i
df_stats_i0 <- func_calculate_change_from_yri_lc0(data = shp_all_df, aoi = aoi_lsoa)
df_chg_sf_i0<- func_calculate_change_from_yri_lc0(data = shp_all_df, aoi = aoi_lsoa, return_sf = T)


# 2. also calculate the change from year 0
df_stats_00 <- func_calculate_change_from_yr0_lc0(data = shp_all_df, aoi = aoi_lsoa)

# 3. and combine the two change calculations  
stat_both_l <- func_combine_two_change_dfs(df_stats_00, df_stats_i0)
```


### Re-code Land scenario

```{r}
## re-code land scenario labels
df_stats_i0 <- df_stats_i0 %>%
  mutate(lc_scenario = factor(lc_scenario, 
                              levels = names(scenario_labels),
                              labels = unname(scenario_labels)))
stat_both_l <- stat_both_l %>%
  mutate(lc_scenario = factor(lc_scenario, 
                              levels = names(scenario_labels),
                              labels = unname(scenario_labels)))

```




## Viz 

  Run either data from the above and then run viz starting from here...
  

```{r - plot - line}

## plot change from year i, old plot 
func_plot_line(data = df_stats_i0 %>% filter(year >= year_baseline))

## plot change from year i 
func_plot_change_point(data = stat_both_l, level = sp_scale, plot_avg = F)
func_plot_change_point(data = stat_both_l, level = sp_scale, plot_h = 4, plot_avg = T)

```


```{r - plot - map}

df_chg_sf_i0_input <- df_chg_sf_i0 %>% 
  filter(year > year_baseline) %>%
  mutate(year = as.factor(year)) %>%
  mutate(lc_scenario = factor(lc_scenario, 
                              levels = names(scenario_labels),
                              labels = unname(scenario_labels)))

## save a copy for viz
f <- paste0('./data/', 'viz_sf_', ind, '_lc_effect_', sp_scale, '.rds'); f
# "./data/viz_sf_avd_eng_cn_lc_effect.rds"
saveRDS(df_chg_sf_i0_input, f)


## apply the function 
func_plot_map(data = df_chg_sf_i0_input, level = sp_scale)
```

 # -----------------------------------------------------------------------------
 #                                      STOP HERE
 # -----------------------------------------------------------------------------


 ### Heavy work productivity loss [OLD model]

  **!!!** please refer to `invest_result_zonal_viz_pd_NEW.Rmd`

  `avg_hvls_v` - (optional) Heavy work productivity loss (%).

  - data stats
   
```{r eval=FALSE, include=FALSE}

ind <- 'avg_hvls_v'
ind_scale <- -1  ## reverse for easier the interpretation: "productivity loss" --> "productivity gain"
ind_label <- "heavy work productivity (%)"


# 1. calculate the change from year i
df_stats_i0 <- func_calculate_change_from_yri_lc0(data = shp_all_df)
df_chg_sf_i0<- func_calculate_change_from_yri_lc0(data = shp_all_df, return_sf = T)


# 2. also calculate the change from year 0
df_stats_00 <- func_calculate_change_from_yr0_lc0(data = shp_all_df)

# 3. and combine the two change calculations  
stat_both_l <- func_combine_two_change_dfs(df_stats_00, df_stats_i0)

```
  
  
  
 - plot - line
 
```{r eval=FALSE, include=FALSE}

## plot change from year i, old plot 
func_plot_line(data = df_stats_i0 %>% filter(year >= year_baseline))

## plot change from year i 
func_plot_change_point(data = stat_both_l, plot_avg = F)
func_plot_change_point(data = stat_both_l, plot_avg = T)

```


 - plot - map
 
```{r eval=FALSE, include=FALSE}

df_chg_sf_i0_input <- df_chg_sf_i0 %>% 
  filter(year > year_baseline) %>%
  mutate(year = as.factor(year)) 


## apply the function 
func_plot_map(data = df_chg_sf_i0_input)

```
