---
title: "Untitled"
author: "Yingjie"
date: "`r Sys.Date()`"
output: html_document

---

# Set up

## Packages 

```{r include=FALSE}

## To clear your environment
remove(list = ls())


## Packages
library(dplyr)
library(tidyr)
library(stringr)
library(sf)

library(ggplot2)
```


## Set Working Directory

> 🔧 **Tip:** Before running this R Markdown file, set the **Knit Directory** to ensure file paths work correctly.

1.  Locate the **Knit** button at the top-left of this panel.
2.  Click the dropdown arrow next to it.
3.  Select **"Knit Directory" → "Project Directory"**.

This ensures that all file paths are relative to your project folder.



# Load data

```
  cur   current/baseline climate 
  fut   future climate (2050 or 2070) 
  MAT   Mean annual temperature 
  MTWM  Maximum temperature of the *warmest* month
  MTCM  Minimum temperature of the *coldest* month 
  AP    Annual precipitation
  PDQ   Precipitation of the driest quarter
```


## Climate data

```{r}

## load data
f <- paste0('./data/', 'spe_at_risk_unique_london.rds')

data1_climate <- readRDS(f)

unique(data1_climate$year)
year_fut <- unique(data1_climate$year)

```



## London tree data

```{r eval=FALSE, include=FALSE}

dir_tree <- "G:/Shared drives/Wellcome Trust Project Data/0_source_data/GiGL land use data/GiGL_Trees_font_point/"

f <- paste0(dir_tree, 'GiGL_GLATrees_Pre2023.shp')

tree <- st_read(f, quiet = TRUE)

tree_sub <- tree %>%
  select(-c(SortOrder, RecMonSta:StatusLISI, SurveyName:SurveyRef, PrivateLoc, PrivateCod,
            RecOccKey:VersionDt))


f <- paste0('./data/', 'GiGL_GLATrees_Pre2023_sub.rds')
saveRDS(object = tree_sub, file = f)
```



```{r load data subset}

f <- paste0('./data/', 'GiGL_GLATrees_Pre2023_sub.rds')
tree_sub <- readRDS(f)

names(tree_sub)

data2_tree_aoi <- tree_sub %>%
  distinct(TaxonName, CommonName, TaxonRank, TaxonGroup, TaxonCat, CommonGrp, TaxonClass, TaxonOrder, TaxonFamil, GroupOrder) %>%
  st_drop_geometry()
```


```{r  tree stats}

## total number of trees in London 
tree_sub %>%
  nrow() %>%
  cat('London has', ., 'trees.\n')

## number of tree species in London 
data2_tree_aoi %>%
  distinct(TaxonName, CommonName, TaxonFamil) %>%
  nrow() %>%
  cat('London has', ., 'tree species in total.\n')


## number of tree at risk 
data1_climate %>%
  distinct(species, family, year) %>%
  nrow() %>%
  cat('London has', ., 'tree species will be at risk.\n')
```

## Identify at-risk trees by name matching

```{r taxize - name matching}
# install.packages("remotes")
# remotes::install_github("ropensci/taxize")


library("taxize")
# resolved <- gnr_resolve(names = c("Quercus robur", "Pinus sylvestris"))



## Match the two dataset: 
# --> data1_climate
# --> data2_tree_aoi


## merge two data - a test 
tree_com1 <- left_join(
  data2_tree_aoi, 
  data1_climate,
  by = c('TaxonName' = 'species')
)



## matched data 1
tree_com1_matched <- tree_com1 %>%
  filter(!is.na(city))


nrow(tree_com1) - nrow(tree_com1_matched)



## for unmatched ones -- find unified name for further matching 

## for data 1 --- 
tree1_0 <- data2_tree_aoi %>%
  filter(!TaxonName %in% unique(tree_com1_matched$TaxonName) ) %>%
  separate(col = TaxonName, into = c("TaxonName1", "TaxonName2"), sep = "=", remove = F) %>%
  mutate(TaxonName1 = trimws(TaxonName1),
         TaxonName2 = trimws(TaxonName2))

tree1_1 <- gna_verifier(tree1_0$TaxonName1)

tree1_1_matched <- tree1_1 %>%
  filter(!is.na(matchedName))

tree1_1_tbd <- tree1_0 %>%
  filter(!TaxonName1 %in% unique(tree1_1_matched$submittedName)) %>%
  mutate(TaxonName3 = paste0(TaxonName1, ' (', TaxonName2, ')'),
         TaxonName3 = gsub('\\(NA\\)', '', TaxonName3)) %>%
  select(1:TaxonName2, TaxonName3, everything())

tree1_2 <- gna_verifier(tree1_1_tbd$TaxonName2)
tree1_3 <- gna_verifier(tree1_1_tbd$CommonName)
tree1_4 <- gna_verifier(tree1_1_tbd$TaxonName3)


## for data 2 ---

tree2_0 <- data1_climate %>%
  filter(!species %in% unique(tree_com1$TaxonName) ) 

tree2_1 <- gna_verifier(tree2_0$species)
tree2_1_matched <- tree2_1 %>%
  filter(!is.na(matchedName)) %>%
  select(submittedName, matchedNameID, matchedName, currentNameId)


## merge again 

tree_com2 <- left_join(
  tree1_1_matched, 
  tree2_1_matched,
  by = 'matchedNameID', ## 0 match
)


tree_com2_matched <- tree_com2 %>%
  filter(!is.na(matchedName.y))
```


```{r final list}

## final data
tree_risk_london <- rbind(tree_com1_matched, tree_com2_matched)

f <- paste0("./data/", "tree_list_GiGL_Pre2023_", year_fut, ".rds")
f <- paste0("./data/", "tree_list_GiGL_Pre2023_", year_fut, ".csv")

readr::write_csv(x = tree_risk_london, f)


## stats
tree_risk_london %>%
  distinct(TaxonName, family, year) %>%
  nrow() %>%
  cat('London has', ., 'tree species will be at risk.\n')
```
