---
title: "Untitled"
author: "Yingjie"
date: "`r Sys.Date()`"
output: html_document

---

# Set up

## Packages 

```{r include=FALSE}

## To clear your environment
remove(list = ls())


## Packages
library(dplyr)
library(tidyr)
library(stringr)
library(sf)

library(ggplot2)
```


## Set Working Directory

> 🔧 **Tip:** Before running this R Markdown file, set the **Knit Directory** to ensure file paths work correctly.

1.  Locate the **Knit** button at the top-left of this panel.
2.  Click the dropdown arrow next to it.
3.  Select **"Knit Directory" → "Project Directory"**.

This ensures that all file paths are relative to your project folder.



# Load data

```
  cur   current/baseline climate 
  fut   future climate (2050 or 2070) 
  MAT   Mean annual temperature 
  MTWM  Maximum temperature of the *warmest* month
  MTCM  Minimum temperature of the *coldest* month 
  AP    Annual precipitation
  PDQ   Precipitation of the driest quarter
```


## Climate data

```{r}

## load data
f <- paste0('./data/', 'spe_at_risk_unique_london_2050.rds')

data1_climate <- readRDS(f)

unique(data1_climate$year)
year_fut <- unique(data1_climate$year)

```



## London tree data

```{r}

dir_tree <- "G:/Shared drives/Wellcome Trust Project Data/0_source_data/GiGL land use data/GiGL_Trees_font_point/"

f.raw <- paste0(dir_tree, 'GiGL_GLATrees_Pre2023.shp')
f.raw <- paste0(dir_tree, 'GiGL_GLATrees_2023-24.shp')

fsub <- paste0('./data/', gsub('.shp', '_sub.rds', basename(f.raw))); fsub
```


```{r eval=FALSE, include=FALSE}

tree <- st_read(f, quiet = TRUE)

tree_sub <- tree %>%
  select(-c(SortOrder, RecMonSta:StatusLISI, SurveyName:SurveyRef, PrivateLoc, PrivateCod,
            RecOccKey:VersionDt))


saveRDS(object = tree_sub, file = fsub)
```



```{r load data subset}

tree_sub <- readRDS(fsub)

names(tree_sub)

data2_tree_aoi <- tree_sub %>%
  distinct(TaxonName, CommonName, TaxonRank, TaxonGroup, TaxonCat, CommonGrp, TaxonClass, TaxonOrder, 
           TaxonFamil, GroupOrder, .keep_all = T) %>%
  st_drop_geometry()
```


```{r  tree stats}

## total number of trees in London 
tree_sub %>%
  nrow() %>%
  cat('London has', ., 'trees.\n')

## number of tree species in London 
data2_tree_aoi %>%
  distinct(TaxonName, CommonName, TaxonFamil) %>%
  nrow() %>%
  cat('London has', ., 'tree species in total.\n')


# ## number of tree at risk 
# data1_climate %>%
#   distinct(species, family, year) %>%
#   nrow() %>%
#   cat('London has', ., 'tree species will be at risk.\n')
```

## Identify at-risk trees by name matching

```{r taxize - name matching}
# install.packages("remotes")
# remotes::install_github("ropensci/taxize")


library("taxize")
# resolved <- gnr_resolve(names = c("Quercus robur", "Pinus sylvestris"))



## Match the two dataset: 
# --> data1_climate
# --> data2_tree_aoi


## merge two data - a test 
tree_com1 <- left_join(
  data2_tree_aoi, 
  data1_climate,
  by = c('TaxonName' = 'species')
)



## matched data 1
tree_com1_matched <- tree_com1 %>%
  filter(!is.na(city))


( nrow(tree_com1) - nrow(tree_com1_matched) ) %>%
  cat('There are', ., 'rows are not matched.\n')



## for unmatched ones -- find unified name for further matching --------------------------

## for data 2 --- 
data2_0 <- data2_tree_aoi %>%
  ## filter un-matched rows
  filter(!TaxonName %in% unique(tree_com1_matched$TaxonName) ) %>%
  separate(col = TaxonName, into = c("TaxonName1", "TaxonName2"), sep = "=", remove = F) %>%
  mutate(TaxonName1 = trimws(TaxonName1),
         TaxonName2 = trimws(TaxonName2))

## To get unified name for matching 
data2_1 <- gna_verifier(data2_0$TaxonName1)

### matched rows
data2_1_matched <- data2_1 %>%
  filter(!is.na(matchedName))

### not matched rows
data2_1_tbd <- data2_0 %>%
  filter(!TaxonName1 %in% unique(data2_1_matched$submittedName)) %>%
  mutate(TaxonName3 = paste0(TaxonName1, ' (', TaxonName2, ')'),
         TaxonName3 = gsub('\\(NA\\)', '', TaxonName3)) %>%
  select(1:TaxonName2, TaxonName3, everything())

### further match based on other columns
data2_2 <- gna_verifier(data2_1_tbd$TaxonName2)
data2_3 <- gna_verifier(data2_1_tbd$CommonName)
data2_4 <- gna_verifier(data2_1_tbd$TaxonName3)


## for data 1 ---

data1_0 <- data1_climate %>%
  filter(!species %in% unique(tree_com1$TaxonName) ) 

data1_1 <- gna_verifier(data1_0$species)
data1_1_matched <- data1_1 %>%
  filter(!is.na(matchedName)) %>%
  select(submittedName, matchedNameID, matchedName, currentNameId)


## merge again 

tree_com2 <- left_join(
  data2_1_matched, 
  data1_1_matched,
  by = 'matchedNameID', ## 0 match
)


tree_com2_matched <- tree_com2 %>%
  filter(!is.na(matchedName.y))


rm(data2_0, data2_1, data2_1_matched, data2_1_tbd, data2_2, data2_3, data2_4,
   data1_0, data1_1, data1_1_matched)
```


```{r final list}
## final data
tree_risk_london <- rbind(tree_com1_matched, tree_com2_matched)

f <- gsub(".shp|GLATrees_", "", basename(f.raw)) %>%
  paste0("./data/", "tree_list_", ., "_", year_fut, ".csv")
f

readr::write_csv(x = tree_risk_london, f)

```


```{r combine data}

## load pre-2023 data
f <- paste0('./data/', 'tree_list_GiGL_pre2023_2050.csv')
tree_risk_london_pre2023 <- readr::read_csv(f, show_col_types = FALSE)


## load 2023-2024 data
f <- paste0('./data/', 'tree_list_GiGL_2023-24_2050.csv')
tree_risk_london_2023 <- readr::read_csv(f, show_col_types = FALSE)


## combine the list
tree_risk_london_combined <- rbind(tree_risk_london_pre2023, tree_risk_london_2023) %>%
  distinct(TaxonName, family, year, country, city, .keep_all = TRUE) %>%
  as.data.frame()

f <- paste0('./data/', 'tree_list_GiGL_combined_2050.csv')
readr::write_csv(x = tree_risk_london_combined, f)

## stats
tree_risk_london_combined %>%
  distinct(TaxonName, family, year) %>%
  nrow() %>%
  cat("London has", ., "tree species will be at risk.\n")
```

