---
title: "Untitled"
author: "Yingjie"
date: "`r Sys.Date()`"
output: html_document

---

# Set up

## Packages 

```{r include=FALSE}

## To clear your environment
remove(list = ls())


## Packages
library(dplyr)
library(tidyr)
library(stringr)
library(sf)

library(ggplot2)
```


## Set Working Directory

> 🔧 **Tip:** Before running this R Markdown file, set the **Knit Directory** to ensure file paths work correctly.

1.  Locate the **Knit** button at the top-left of this panel.
2.  Click the dropdown arrow next to it.
3.  Select **"Knit Directory" → "Project Directory"**.

This ensures that all file paths are relative to your project folder.



# Load data

cur current/baseline climate 
fut future climate (2050 or 2070) 
MAT Mean annual temperature 
MTWM Maximum temperature of the *warmest* month
MTCM Minimum temperature of the *coldest* month 
AP Annual precipitation
PDQ Precipitation of the driest quarter


## Global data

### temp data for London

```{r}

dir <- "G:/Shared drives/Wellcome Trust Project Data/0_source_data/Climate-urban-forests/"

f <- paste0(dir, "2_cities_climate_exposure.csv")

city_climate <- readr::read_csv(f, show_col_types = F)

c_climate <- city_climate %>%
  filter(city == "London") %>%
  select(cur_MTWM, fut_MTWM_2050_45, fut_MTWM_2070_45) %>%
  gather(key = "var", value = "value") %>%
  mutate(var = gsub("_MTWM|_45|fut_MTWM_", "", var)) %>%
  mutate(var = ifelse(var == "cur", "Baseline", var)) %>%
  mutate(var = factor(x = var, levels = c("Baseline", "2050", "2070")))



## plot
c_climate %>%
  ggplot(aes(x = var, y = value, fill = value)) +
  geom_col(show.legend = F) +
  labs(
    x = NULL, y = expression("Temperature (" * ~ degree * C * ")"),
    title = "Max temperature of the warmest month \nin London, UK"
  ) +
  # scale_fill_distiller(palette = "YlOrBr", direction = 1) +
  scale_fill_gradient(low = "#fed98e", high = "#d95f0e") + # Lighter to moderate orange
  coord_cartesian(ylim = c(20, 25)) +
  geom_text(aes(label = round(value, 1)), vjust = -0.5, size = 4) + # Add value labels
  theme_minimal(base_size = 18) +
  theme(plot.title = element_text(hjust = 0.5))

f <- "./figures/temp_scenario.png"
ggsave(f, plot = last_plot(), width = 6, height = 7, units = "in", dpi = 300)
```


### species at risk

If the exposure to future climate is greater than the current safety margin for the focal species in a focal city (that is, *high risk*), then 
  * R > 0 for MAT and MTWM and 
  * R < 0 for MTCM, AP and PDQ.

```{r}

f <- paste0(dir, "5_results_complete.csv")

year_fut <- 2050
year_fut <- 2070

spe <- readr::read_csv(f, show_col_types = F) 
# names(spe)

spe_climate <- spe %>%
  filter(city == 'London') %>%
  select(1:family, 
         matches("MTWM|safety|risk", ignore.case = F) ) %>%
  select(-matches("MAT_|PDQ|_AP|_MAT|_60"))

# names(spe_climate)


## 
spe_at_risk <- spe %>%
  filter(city == 'London') %>%
  select(1:family, 
         matches("risk", ignore.case = F) ) %>%
  select(-matches("_60")) %>%
  
  filter(
    risk_MAT_95_2050_45 >0   | risk_MAT_95_2070_45 > 0 | 
      risk_MTWM_95_2050_45>0 | risk_MTWM_95_2070_45>0 |
      risk_AP_95_2050_45 < 0 | risk_AP_95_2070_45<0 |
      risk_PDQ_95_2050_45<0  | risk_PDQ_95_2070_45<0
  ) %>%
  gather(key = 'var', value = 'value', 5:ncol(.)) %>%
  mutate(year = str_extract(var, "20(50|70)"),     # extract "2030", "2040", or "2050"
         year = as.numeric(year))  %>%
  filter(year == year_fut)


# unique(spe_at_risk$var)
# names(spe_at_risk)


## get a list of unique tree species that will be at risk
spe_at_risk_unique <- spe_at_risk %>%
  distinct(country, city, species, family, year)


## save data
f <- paste0('./data/', 'spe_at_risk_unique_london_', year_fut, '.rds')
saveRDS(spe_at_risk_unique, f)
```

